Facility and Time Step Discretization Effects in Fuel Cycle Simulation

### 2015-09-30 paper.tex
adjust y range and legend placement for puinv figure
add todos
relabel four cases with MI,MF,QI,QF mneumonics. Add a bit more discussion of striations

@@

    \item \emph{Case 1}: Monthly time steps with individual reactor modeling
    \item \emph{Case 2}: Monthly time steps with fleet reactor modeling
    \item \emph{Case 3}: Quarterly (3-month) time steps with individual reactor modeling
    \item \emph{Case 4}: Quarterly (3-month) time steps with fleet reactor modeling

@@

A pre-release of Cyclus version 1.4 was used for this analysis.  Archetypes
used were the standard ones provided with Cyclus in the included Cycamore
library - also a pre-release of version 1.4.  A custom storage facility was
used because Cycamore did not yet provide one.  The standard cycamore reactor
was used for both LWR and SRF reactor types in cases 1 and 3.  Correspondingly
the custom fleet-based reactor described earlier was used for both reactor
types in cases 2 and 4.  Additional configuration for the reactor types is
shown in Table \ref{tab:reactors}.

@@

\begin{table*}
    \centering
    \begin{tabular}{ |r | c c | c c | }
        \hline                       
                                          & \multicolumn{2}{c |}{LWR}       & \multicolumn{2}{c |}{SFR} \\
                                          & Cases 1,3 & Cases 2,4 (fleet) & Cases 1,3 & Cases 2,4 (fleet)  \\
        \hline                       
        Cycle length (months)             & 15        & 18                & 12        & 15 \\
        Refueling outage (months)         & 3         & 0                 & 3         & 0 \\
        Batch size ($\text{kg} \cdot \text{HM}$) & 29,565    & 29,565            & 8,025     & 40,125 \\
        Power capacity (MWe)              & 1,080     & 900               & 450       & 360 \\
        \hline                       
    \end{tabular}
    \captionsetup{justification=centering}
    \caption[Reactor parameters by case]{
        Reactor configuration for each of the four simulation cases.
    }

@@

One difference between the fleet and individual facility modeling is the
\emph{quantized shutdown effect}.  If a fleet-based reactor modeling 3-batch
cores is short one half batch of fuel, it will produce $\frac{2.5}{3}\cdot C$
of power for that time step where $C$ is the power capacity of a single
reactor of the fleet (e.g. 1000 MWe).  If an individually modeled reactor is
short a half batch of fuel, then it will produce no power ($0 \cdot C$) that
time step.  Time step duration can also have this effect.  For both fleet and
individual reactor modeling, power goes offline in full time step increments.
For example, if a 3-batch fleet based reactor is short $\frac{1}{2}$ batch of
fuel with a 3 month time step, then it will produce $\frac{2.5}{3} \cdot 3
\text{mo} \cdot C $ of energy.  In contrast, a monthly time step might allow
the same reactor to produce $[\frac{5}{6} \cdot 1 \text{mo} + 1 \cdot 2
\text{mo}] \cdot C $ of energy if the fuel shortage actually ended after only
1 month.  Quantifying this effect directly is difficult.  One way of observing
it is to compare fleet reactor scenarios that have different time step
durations (i.e. cases 2 and 4).  Because the fleet reactors have perfect fuel
sharing (described below) and no outage noise, differences observed will
likely be due to quantized shutdown.

@@

One important phenomenon related to time step duration is the \emph{inventory
drawdown effect}.  Increasing time step duration reduces the frequency over
which refueling can occur, resulting in larger impulse drawdown on
inventories. This by itself is not problematic because this is balanced by
correspondingly larger inventory top-up quantities.  The other component of
this effect is the simultaneity of incoming and outgoing material flows on a
given time step.  Within a particular time step facilities do not know about
incoming material when they make commitments for outgoing material.  For
longer time steps, larger incoming quanta of material are not available for
making offers. In general longer time steps create a need for larger floating
inventory buffers in order to avoid material shortages.  Another aspect of
this effect is that a single material object can only move between a single
pair of facilities per time step. Larger time steps result in longer minimum
bounds on time to traverse paths between facilities. Although direct measures
of this effect are difficult, some of its consequences (e.g. higher standing
inventories during shortages) can be quantified by comparing scenarios with
different time step durations (e.g. cases 2 and 4).

@@

Figure \ref{fig:power} shows an overall view of the generated power over time
for each of the four cases.  Figure \ref{fig:power-rel} normalizes the Figure
\ref{fig:power} curves to the expected exponential power curve magnifying some
interesting differences between the four cases.  For the first 100 years, the
four cases behave somewhat similarly, although the individually modeled
reactors in cases 1 and 3 actually go offline and back online for refueling
causing more variance.  Around year 100, a fuel shortage begins and more
significant differences between the four cases become apparent. This fuel
shortage lasts until about year 140. The power generated in Case 4 is slightly
lower than in case 2 during the fuel shortage years. This is primarily due to
the \emph{quantized shutdown effect}.  The longer 3-month time step in case 4
results in reactor capacity going offline longer than necessary.

During the initial 100 years, cases 1 and 3 have larger variance in power
output than cases 2 and 4.  This is caused by minor refueling cycle
synchronization.  The \emph{cycle staggering effect} becomes much more visible
during and after the fuel shortage as evidenced by the larger swings in power
generation from time step to time step.  Initially, the reactor cycles are
staggered well.  During the shortage, several reactors that previously had
staggered cycles are all offline together waiting for fuel.  On time steps
where reactors retire (not just during the shortage), new deployments are made
to replace them in addition to new deployments made to address power demand
growth.  Because the initial reactors retire in waves somewhat close together,
there are corresponding waves of deployment, and these waves are echoed every
80 years (the reactor lifetime).  These waves of deployment cause waves in
recycled fuel availability that cause many of the reactors offline during the
shortage to receive fuel and come online together.  The net effect is that the
fuel shortage degrades reactor cycle stagering overall.  As the simulation
continues, however, new reactors continue to be deployed and reactors with
synchornized cycles retire resulting in a gradual return to decent better
staggering visible by the end of the 200 years.

\begin{figure*}[!h]
    \centering
    \includegraphics[width=0.8\textwidth]{exp2/power.eps}
    \caption[Generated power]{
        Generated power for all four cases.  A fuel shortage occurs from about
        year 100 to about year 140.  In cases 1 and 3 with individual reactor
        modeling, refueling cycles become much more synchronized during and
        after the fuel shortage resuting in large power fluctuations - an artifact of the
        \emph{cycle staggering effect}.
    }
    \label{fig:power}
\end{figure*}

\begin{figure*}[!h]
    \centering
    \includegraphics[width=0.8\textwidth]{exp2/power-rel.eps}
    \caption[Normalized power]{
        Generated power is normalized to the expected 1\% exponential growth
        curve for all four cases.  A fuel shortage occurs from about year 100
        to about year 140.  The \emph{cycle staggering effect} can be seen as
        excessive divergence above and below $1.0$ in cases 1 and 3 where
        the shortage increases cycle synchronization significantly. The
        \emph{quantized shutdown effect} can also be seen in the discrepency
        between quarterly and monthly time steps for fleet reactor modeling
        (cases 2 and 4) during the shortage; a longer time step causes more
        reactor outage than necessary. The consistent jumping up and down for
        the fleet reactor cases (even pre-shortage) is due to the mis-alignment of reactors going
        offline (multiples of 12 months) and the build period (21 months).
    }
    \label{fig:power-rel}
\end{figure*}

@@

Figure \ref{fig:unfueled} shows the fuel shortage more explicitly for each of
the four cases. The individual reactor modeling cases result in more
cumulative outage than corresponding fleet-based cases. The quarterly time
step of cases 3 and 4 also make the outage significantly more severe. Cases 1
and 3 exhibit approximately twice as much offline power as their corresponding
fleet based simulations in cases 2 and 4.  The monthly time step simulations
also have roughly twice as much offline power as their corresponding quarterly
cases.

@@

One useful over-estimate of the fuel sharing inefficiency is to assume every
one of the "wasted" batches could have been given to a reactor enabling it to
operate.  Multiplying each of these cumulative batches by reactor capacity
(i.e. 450 MWe) provides a way to compare fuel sharing inefficiency with the
overall fuel outage.  This results in a cumulative offline power caused by
poor fuel sharing of roughly 2,000 $\text{GWe} \cdot \text{months}$ and 6,600
$\text{GWe} \cdot \text{months}$ for cases 1 and 3 respectively.  These
approximations are a significant portion of the difference between individual
and fleet cases in Figure \ref{fig:unfueled} suggesting that most of the
discrepency in cumulative outage between fleet and individual modeling is
caused by inefficient fuel sharing in the individual reactor cases.

@@

If they were to ever occur, real world fuel shortages would likely not result
in optimal fuel sharing. However, modeling these the causes of suboptimal fuel
sharing is probably best accomplished with a more direct, intentional approach
rather than as a modeling artifact as seen here in cases 1 and 3.  One way to
alleviate the poor fuel sharing is to modify the individual reactor model to
lower the preference value for requests for fresh fuel depending on how many
assemblies it needs to have a full core.  The more assemblies it needs, the
lower it will set its request preference.  This will have the effect of
allowing the resource exchange resolution in Cyclus to globally prefer sending
fuel to reactors that need less to operate.

\begin{figure}[!h]
    \centering
    \includegraphics[width=1.0\columnwidth]{exp2/badshare.eps}
    \caption[Cumulative unnecessary idling fuel]{
        Fuel sharing inefficiency approximated by the cumulative number of
        fuel batches given to each reactor on each refueling period multiplied
        by the number of time steps that reactor had to delay the start of its
        next cycle (due to fuel shortage).  Fleet reactors have perfect fuel
        sharing by design and so are not included here. The large difference
        between cases 1 and 3 is mostly due to the \emph{inventory drawdown
        effect}. 
    }
    \label{fig:badshare}
\end{figure}

@@

The large differences seen between monthly and quarterly time step cases in
Figures \ref{fig:unfueled} and \ref{fig:badshare} are primarily caused by the
\emph{inventory drawdown effect}.  Figure \ref{fig:puinv} helps to visualize
this effect.  The black dots in the figure represent impulse flows of
available Pu out of the fuel fab and into fast reactors.  Blue lines represent
the flows into Pu inventory available for making fuel. The in flows are not
shown for the individual reactor cases because they are somewhat messy and
obscure other information in the plots. The other colored curves show Pu inventory
available for making fast reactor fuel.  When the inventory curves meet and
force down the out-flows, fuel shortages occur.  While this is most clear for
cases 2 and 4, the individual reactor scenarios (cases 1 and 3) also have many
points during the shortage where the outflow point lies exactly on top of the
inventory point - indicating that all available inventory is being
transferred.

Quarterly time steps have less frequent but larger impulse material flows.
The larger in-flows, however, do not compensate for the larger withdrawals
because facilities do not know about incoming inventory when they resolve
their outflow for a particular time step.  This information lag effectively
requires the floating Pu inventory to be maintained at a higher level during
the shortage.  This can be clearly seen in the case 4 plot which has a higher
standing inventory than case 2 during the fuel shortage years.  The case 4 Pu
inventory is unable to drop below approximately 100 tonnes, which indicates
out-flows to be roughly equal to the Pu flow into inventory each time step
during those years.  This can be confirmed by the in-flow and inventory curves
touching at those points.  Case 2 Pu inventory is similarly limited at about
the 30 tonne level.  As expected, the case 4 withdrawals are also three times
higher case 2 withdrawals exactly matching the time step duration ratio
between the two cases.

Throughout most of the shortage, the aggregate outflow of Pu from separated
inventory is roughly the same in cases 2 and 4.  However, due to the longer
time steps in case 4, the shortage starts slightly sooner because Pu inventory
must be maintained at a higher limit as described above.  The small difference
in outage start time has feed-back effects.  Not only are extra reactors
offline, but those extra reactors are also discharging less aggregate spent
fuel for recycling.  This farther increases the discrepency between the two
cases. Eventually the shortage in case 2 ends slightly before it does in case
4. This effect feedback occurs because the reactors in question have a
conversion ratio greater than 1.0.

The striations visible in the monthly flows for case 1 in Figure
\ref{fig:puinv} are a characteristic of the synchronization of reactor
refueling caused by the shortage.  They are a sort of resonances of that the
inventory level jumps between, and they gradually disappear as the refueling
cycles become more staggered.

\begin{figure*}[!h]
    \centering
    \includegraphics[width=1.0\textwidth]{exp2/puinv.eps}
    \caption[Separated Pu inventory and flow]{

        Separated Pu inventory and flows for all four cases.  In-flows for
        cases 1 and 3 are omitted because they are noisy and obscure other
        useful information.  A larger time step results in larger per time
        step material flows. These larger flows mean more material is not
        available for supplying to requesters.  This information lag causes
        more supply constraints than occur with smaller time steps -
        illustrating the \emph{inventory drawdown effect}. Another result of
        this effect is the larger standing Pu inventory during the shortage
        with larger time steps that can be seen in case 4 than in case 2.

    }
    \label{fig:puinv}
\end{figure*}

Figure \ref{fig:puinv-compare} shows the same inventory curves from Figure
\ref{fig:puinv} superimposed. Perhaps counter-intuitively, lower separated plutonium
inventory during the shortage generally indicates fewer unfueled reactors.
Lower inventory levels mean available Pu is being utilized more efficiently
rather than sitting idle.  Cases 3 and 4 with their longer time steps show
higher inventories indicating a more severe shortage.  Post shortage, they
also fall a bit behind on building up Pu stocks due to the delayed
contributions of more unfueled reactors to the separated Pu pool.  The
reactors in all the scenarios request fresh fuel only when they need it. While
this is an unrealistic behavior for real-world facilities, it is closer to a
globally optimum fuel management strategy.  Keeping on-hand fresh fuel batches
would also have the same effect of increasing idling plutonium quantities and
would make the fuel shortage worse overall.

>>>>

TODO: add table titles
TODO: adjust figure fonts to meet NT guidlines

@@

    \item \emph{Case MI}: Monthly time steps with individual reactor modeling
    \item \emph{Case MF}: Monthly time steps with fleet reactor modeling
    \item \emph{Case QI}: Quarterly (3-month) time steps with individual reactor modeling
    \item \emph{Case QF}: Quarterly (3-month) time steps with fleet reactor modeling

The case labels consist of two letters each: "M" for monthly time steps, "Q"
for quarterly time steps, "I" for individual reactors, and "F" for fleet
reactors.

@@

A pre-release of Cyclus version 1.4 was used for this analysis.  Archetypes
used were the standard ones provided with Cyclus in the included Cycamore
library - also a pre-release of version 1.4.  A custom storage facility was
used because Cycamore did not yet provide one.  The standard cycamore reactor
was used for both LWR and SRF reactor types in cases MI and QI.  Correspondingly
the custom fleet-based reactor described earlier was used for both reactor
types in cases MF and QF.  Additional configuration for the reactor types is
shown in Table \ref{tab:reactors}.

@@

\begin{table*}
    \centering
    \begin{tabular}{ |r | c c | c c | }
        \hline                       
                                          & \multicolumn{2}{c |}{LWR}       & \multicolumn{2}{c |}{SFR} \\
                                          & Cases MI,QI & Cases MF,QF (fleet) & Cases MI,QI & Cases MF,QF (fleet)  \\
        \hline                       
        Cycle length (months)             & 15        & 18                & 12        & 15 \\
        Refueling outage (months)         & 3         & 0                 & 3         & 0 \\
        Batch size ($\text{kg} \cdot \text{HM}$) & 29,565    & 29,565            & 8,025     & 40,125 \\
        Power capacity (MWe)              & 1,080     & 900               & 450       & 360 \\
        \hline                       
    \end{tabular}
    \captionsetup{justification=centering}
    \caption[Reactor parameters by case]{
        Reactor configuration for each of the four simulation cases.
    }

@@

One difference between the fleet and individual facility modeling is the
\emph{quantized shutdown effect}.  If a fleet-based reactor modeling 3-batch
cores is short one half batch of fuel, it will produce $\frac{2.5}{3}\cdot C$
of power for that time step where $C$ is the power capacity of a single
reactor of the fleet (e.g. 1000 MWe).  If an individually modeled reactor is
short a half batch of fuel, then it will produce no power ($0 \cdot C$) that
time step.  Time step duration can also have this effect.  For both fleet and
individual reactor modeling, power goes offline in full time step increments.
For example, if a 3-batch fleet based reactor is short $\frac{1}{2}$ batch of
fuel with a 3 month time step, then it will produce $\frac{2.5}{3} \cdot 3
\text{mo} \cdot C $ of energy.  In contrast, a monthly time step might allow
the same reactor to produce $[\frac{5}{6} \cdot 1 \text{mo} + 1 \cdot 2
\text{mo}] \cdot C $ of energy if the fuel shortage actually ended after only
1 month.  Quantifying this effect directly is difficult.  One way of observing
it is to compare fleet reactor scenarios that have different time step
durations (i.e. cases MF and QF).  Because the fleet reactors have perfect fuel
sharing (described below) and no outage noise, differences observed will
likely be due to quantized shutdown.

@@

One important phenomenon related to time step duration is the \emph{inventory
drawdown effect}.  Increasing time step duration reduces the frequency over
which refueling can occur, resulting in larger impulse drawdown on
inventories. This by itself is not problematic because this is balanced by
correspondingly larger inventory top-up quantities.  The other component of
this effect is the simultaneity of incoming and outgoing material flows on a
given time step.  Within a particular time step facilities do not know about
incoming material when they make commitments for outgoing material.  For
longer time steps, larger incoming quanta of material are not available for
making offers. In general longer time steps create a need for larger floating
inventory buffers in order to avoid material shortages.  Another aspect of
this effect is that a single material object can only move between a single
pair of facilities per time step. Larger time steps result in longer minimum
bounds on time to traverse paths between facilities. Although direct measures
of this effect are difficult, some of its consequences (e.g. higher standing
inventories during shortages) can be quantified by comparing scenarios with
different time step durations (e.g. cases MF and QF).

@@

Figure \ref{fig:power} shows an overall view of the generated power over time
for each of the four cases.  Figure \ref{fig:power-rel} normalizes the Figure
\ref{fig:power} curves to the expected exponential power curve magnifying some
interesting differences between the four cases.  For the first 100 years, the
four cases behave somewhat similarly, although the individually modeled
reactors in cases MI and QI actually go offline and back online for refueling
causing more variance.  Around year 100, a fuel shortage begins and more
significant differences between the four cases become apparent. This fuel
shortage lasts until about year 140. The power generated in case QF is slightly
lower than in case MF during the fuel shortage years. This is primarily due to
the \emph{quantized shutdown effect}.  The longer 3-month time step in case QF 
results in reactor capacity going offline longer than necessary.

During the initial 100 years, cases MI and QI have larger variance in power
output than cases MF and QF.  This is caused by minor refueling cycle
synchronization.  The \emph{cycle staggering effect} becomes much more visible
during and after the fuel shortage as evidenced by the larger swings in power
generation from time step to time step.  Initially, the reactor cycles are
staggered well.  During the shortage, several reactors that previously had
staggered cycles are all offline together waiting for fuel.  On time steps
where reactors retire (not just during the shortage), new deployments are made
to replace them in addition to new deployments made to address power demand
growth.  Because the initial reactors retire in waves somewhat close together,
there are corresponding waves of deployment, and these waves are echoed every
80 years (the reactor lifetime).  These waves of deployment cause waves in
recycled fuel availability that cause many of the reactors offline during the
shortage to receive fuel and come online together.  The net effect is that the
fuel shortage degrades reactor cycle stagering overall.  As the simulation
continues, however, new reactors continue to be deployed and reactors with
synchornized cycles retire resulting in a gradual return to decent better
staggering visible by the end of the 200 years.

\begin{figure*}[!h]
    \centering
    \includegraphics[width=0.8\textwidth]{exp2/power.eps}
    \caption[Generated power]{
        Generated power for all four cases.  A fuel shortage occurs from about
        year 100 to about year 140.  In cases MI and QI with individual reactor
        modeling, refueling cycles become much more synchronized during and
        after the fuel shortage resuting in large power fluctuations - an artifact of the
        \emph{cycle staggering effect}.
    }
    \label{fig:power}
\end{figure*}

\begin{figure*}[!h]
    \centering
    \includegraphics[width=0.8\textwidth]{exp2/power-rel.eps}
    \caption[Normalized power]{
        Generated power is normalized to the expected 1\% exponential growth
        curve for all four cases.  A fuel shortage occurs from about year 100
        to about year 140.  The \emph{cycle staggering effect} can be seen as
        excessive divergence above and below $1.0$ in cases MI and QI where
        the shortage increases cycle synchronization significantly. The
        \emph{quantized shutdown effect} can also be seen in the discrepency
        between quarterly and monthly time steps for fleet reactor modeling
        (cases MF and QF) during the shortage; a longer time step causes more
        reactor outage than necessary. The consistent jumping up and down for
        the fleet reactor cases (even pre-shortage) is due to the mis-alignment of reactors going
        offline (multiples of 12 months) and the build period (21 months).
    }
    \label{fig:power-rel}
\end{figure*}

@@

Figure \ref{fig:unfueled} shows the fuel shortage more explicitly for each of
the four cases. The individual reactor modeling cases result in more
cumulative outage than corresponding fleet-based cases. The quarterly time
step of cases QI and QF also make the outage significantly more severe. Cases MI
and QI exhibit approximately twice as much offline power as their corresponding
fleet based simulations in cases MF and QF.  The monthly time step simulations
also have roughly twice as much offline power as their corresponding quarterly
cases.

@@

One useful over-estimate of the fuel sharing inefficiency is to assume every
one of the "wasted" batches could have been given to a reactor enabling it to
operate.  Multiplying each of these cumulative batches by reactor capacity
(i.e. 450 MWe) provides a way to compare fuel sharing inefficiency with the
overall fuel outage.  This results in a cumulative offline power caused by
poor fuel sharing of roughly 2,000 $\text{GWe} \cdot \text{months}$ and 6,600
$\text{GWe} \cdot \text{months}$ for cases MI and QI respectively.  These
approximations are a significant portion of the difference between individual
and fleet cases in Figure \ref{fig:unfueled} suggesting that most of the
discrepency in cumulative outage between fleet and individual modeling is
caused by inefficient fuel sharing in the individual reactor cases.

@@

If they were to ever occur, real world fuel shortages would likely not result
in optimal fuel sharing. However, modeling these the causes of suboptimal fuel
sharing is probably best accomplished with a more direct, intentional approach
rather than as a modeling artifact as seen here in cases MI and QI.  One way to
alleviate the poor fuel sharing is to modify the individual reactor model to
lower the preference value for requests for fresh fuel depending on how many
assemblies it needs to have a full core.  The more assemblies it needs, the
lower it will set its request preference.  This will have the effect of
allowing the resource exchange resolution in Cyclus to globally prefer sending
fuel to reactors that need less to operate.

\begin{figure}[!h]
    \centering
    \includegraphics[width=1.0\columnwidth]{exp2/badshare.eps}
    \caption[Cumulative unnecessary idling fuel]{
        Fuel sharing inefficiency approximated by the cumulative number of
        fuel batches given to each reactor on each refueling period multiplied
        by the number of time steps that reactor had to delay the start of its
        next cycle (due to fuel shortage).  Fleet reactors have perfect fuel
        sharing by design and so are not included here. The large difference
        between cases MI and QI is mostly due to the \emph{inventory drawdown
        effect}. 
    }
    \label{fig:badshare}
\end{figure}

@@

The large differences seen between monthly and quarterly time step cases in
Figures \ref{fig:unfueled} and \ref{fig:badshare} are primarily caused by the
\emph{inventory drawdown effect}.  Figure \ref{fig:puinv} helps to visualize
this effect.  The black dots in the figure represent impulse flows of
available Pu out of the fuel fab and into fast reactors.  Blue lines represent
the flows into Pu inventory available for making fuel. The in flows are not
shown for the individual reactor cases because they are somewhat messy and
obscure other information in the plots. The other colored curves show Pu inventory
available for making fast reactor fuel.  When the inventory curves meet and
force down the out-flows, fuel shortages occur.  While this is most clear for
cases MF and QF, the individual reactor scenarios (cases MI and QI) also have many
points during the shortage where the outflow point lies exactly on top of the
inventory point - indicating that all available inventory is being
transferred.

Quarterly time steps have less frequent but larger impulse material flows.
The larger in-flows, however, do not compensate for the larger withdrawals
because facilities do not know about incoming inventory when they resolve
their outflow for a particular time step.  This information lag effectively
requires the floating Pu inventory to be maintained at a higher level during
the shortage.  This can be clearly seen in the case QF plot which has a higher
standing inventory than case MF during the fuel shortage years.  The case QF Pu
inventory is unable to drop below approximately 100 tonnes, which indicates
out-flows to be roughly equal to the Pu flow into inventory each time step
during those years.  This can be confirmed by the in-flow and inventory curves
touching at those points.  Case MF Pu inventory is similarly limited at about
the 30 tonne level.  As expected, the case QF withdrawals are also three times
higher case MF withdrawals exactly matching the time step duration ratio
between the two cases.

Throughout most of the shortage, the aggregate outflow of Pu from separated
inventory is roughly the same in cases MF and QF.  However, due to the longer
time steps in case QF, the shortage starts slightly sooner because Pu inventory
must be maintained at a higher limit as described above.  The small difference
in outage start time has feed-back effects.  Not only are extra reactors
offline, but those extra reactors are also discharging less aggregate spent
fuel for recycling.  This farther increases the discrepency between the two
cases. Eventually the shortage in case MF ends slightly before it does in case
QF. This effect feedback occurs because the reactors in question have a
conversion ratio greater than 1.0.

The striations visible in the monthly flows for case MI in Figure
\ref{fig:puinv} are a characteristic of the synchronization of reactor
refueling caused by the shortage.  They begin at the end of the fuel shortage
window when many reactors come online together and are resonances that the
inventory level jumps between.  These resonances gradually disappear as the
simulation progresses farther in time beyond the shortage and refueling cycles
become more staggered.

\begin{figure*}[!h]
    \centering
    \includegraphics[width=1.0\textwidth]{exp2/puinv.eps}
    \caption[Separated Pu inventory and flow]{
        Separated Pu inventory and flows for all four cases.  In-flows for
        cases MI and QI are omitted because they are noisy and obscure other
        useful information.  A larger time step results in larger per time
        step material flows. These larger flows mean more material is not
        available for supplying to requesters.  This information lag causes
        more supply constraints than occur with smaller time steps -
        illustrating the \emph{inventory drawdown effect}. Another result of
        this effect is the larger standing Pu inventory during the shortage
        with larger time steps that can be seen in case QF than in case MF.
    }
    \label{fig:puinv}
\end{figure*}

Figure \ref{fig:puinv-compare} shows the same inventory curves from Figure
\ref{fig:puinv} superimposed. Perhaps counter-intuitively, lower separated plutonium
inventory during the shortage generally indicates fewer unfueled reactors.
Lower inventory levels mean available Pu is being utilized more efficiently
rather than sitting idle.  Cases QI and QF with their longer time steps show
higher inventories indicating a more severe shortage.  Post shortage, they
also fall a bit behind on building up Pu stocks due to the delayed
contributions of more unfueled reactors to the separated Pu pool.  The
reactors in all the scenarios request fresh fuel only when they need it. While
this is an unrealistic behavior for real-world facilities, it is closer to a
globally optimum fuel management strategy.  Keeping on-hand fresh fuel batches
would also have the same effect of increasing idling plutonium quantities and
would make the fuel shortage worse overall.
